---
title: "RSTooDs"
subtitle: "An R user interface to STooDs"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,results='hide',
                      fig.path = "man/readme/README-")
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5075760.svg)](https://doi.org/10.5281/zenodo.5075760)

# Introduction

STooDs is a framework to build and estimate probabilistic models for data varying in Space, Time or other Dimensions. The R package `RSTooDs` is built as an R User Interface to the [computational STooDs engine](https://github.com/STooDs-tools/STooDs). It defines classes (objects' properties and methods) for the various building blocks of a STooDs case study. Its typical usage is as follows:

1. Assemble the dataset.
2. Define the probabilistic model and its constitutive objects (parameters, dimensions, processes).
3. Perform Bayesian-MCMC inference.
4. Read, analyse and use MCMC samples.

```{r}
library(RSTooDs)
```

**Important warning**: many distributions are available in `RSTooDs`, but for some of them the parameterization may differ from the one used in e.g. Wikipedia or other R packages. The dataset `distInfo` provides information on the distributions *as they are used in* `RSTooDs`: it is highly advised to read this information before using a distribution. Try the following:

```{r,results=FALSE,message=FALSE}
# Show available distributions
names(distInfo)
# Show information on e.g. the Generalized Extreme Value (GEV) distribution
distInfo[['GEV']]
# Show all existing warnings
showWarnings()
```

# A simple example

River streamflow in Eastern Australia is influenced by the [El Ni単o Southern Oscillation](https://en.wikipedia.org/wiki/El_Ni単o-Southern_Oscillation), as explained in this [blog post](https://globxblog.inrae.fr/el-ninoz/). As an illustration, let's consider data from the Barnard River in New South Wales, Australia. The figures below show the average streamflow during the austral spring (September to November), and indeed it seems that negative values of the [nino3.4 index](https://psl.noaa.gov/gcos_wgsp/Timeseries/Nino34/), corresponding to [La Ni単a episodes](https://en.wikipedia.org/wiki/La_Ni単a), are associated with large streamflow.

```{r fig.height=8, fig.width=9}
dat=read.table(file.path('man','readme','BarnardRiverStreamflow.txt'),header=TRUE)
par(mfrow=c(2,1))
plot(dat$year,dat$streamflow,type='b',pch=19,xlab='Year',ylab='Streamflow (m3/s)',main='Time series')
plot(dat$nino,dat$streamflow,type='p',pch=19,xlab='Nino index',ylab='Streamflow (m3/s)',main='Association with Nino3.4 index')
```

A probabilistic model can be used to quantify the association between river streamflow and the nino index. For instance one could assume that observed streamflow values are realizations from a log-normal distribution whose first parameter (mu) varies as a function of the nino index. The few lines of codes below show how this can be specified with `RSTooDs`.

```{r}
# Assemble the dataset with predictand Y and predictor X (aka covariate).
D=dataset(Y=dat['streamflow'],X=dat['nino'])
# Define formulas of the model
f=c('mu=m0+m1*nino','sigma=s0')
# Define parameters of the model and their prior distributions
m0=parameter('m0',init=1) # Improper flat prior by default
m1=parameter('m1',init=0,priorDist='Gaussian',priorPar=c(0,1))
s0=parameter('s0',init=1,priorDist='FlatPrior+')
# Assemble the model - type getCatalogue() for a list of available distributions.
mod=model(dataset=D,parentDist='LogNormal',par=list(m0,m1,s0),formula=f)
```

Once the model is specified, the function `STooDs` can be called to perform MCMC sampling. It is recommended to first define a `workspace` folder where configuration and result files will be written.

```{r}
wk=file.path(getwd(),'man','readme','wk')
STooDs(model=mod,workspace=wk)
```

MCMC samples can then be explored as illustrated below. Traces in orange correspond to the parameters, traces in blue correspond to the standard Bayesian inference functions (all in log space): prior, likelihood, hierarchical component (not used here) and posterior.

```{r fig.height=3, fig.width=9}
mcmc=readMCMC(file=file.path(wk,'MCMC.txt'))
plotMCMC.trace(mcmc,mod,panelPerCol=4)
```

The function below plots histograms of all parameters. The parameter m1 appears to be largely negative, corresponding to the observed negative association between streamflow and the nino index.

```{r fig.height=3, fig.width=9}
plotMCMC.par(mcmc,mod)
```

# A more complex example using spatial processes

It is legitimate to ask whether the effect of El Nino would be the same, or would at least be similar, for nearby rivers. The data shown below correspond to 21 stations located in New South Wales and Queensland (source: [Bureau of Meteorology](http://www.bom.gov.au/water/hrs/)) and can be used to investigate this question.

```{r fig.height=8, fig.width=9}
dat=read.table(file.path('man','readme','21RiversStreamflow.txt'),header=TRUE)
stations=read.table(file.path('man','readme','21Rivers.txt'),header=TRUE)
par(mfrow=c(2,1))
plot(dat$year,dat$streamflow,type='p',pch=19,col=dat$space_index,xlab='Year',ylab='Streamflow (m3/s)',log='y',main='Time series')
plot(dat$nino,dat$streamflow,type='p',pch=19,col=dat$space_index,xlab='Nino index',ylab='Streamflow (m3/s)',log='y',main='Association with Nino3.4 index')
```

We are going to use a model similar to the previous one: observed streamflows are realizations from a log-normal distribution whose first parameter (mu) varies in time as a function of the nino index. However, since data are now varying in space as well, it is necessary to modify the model so that parameters may also vary in space. This can be achieved by first defining a space `dimension`, and then by attaching `processes` to it. In a nutshell, a process can hence be viewed as a parameter that varies along a given dimension.

```{r}
# Assemble the dataset. Note the use of iDim (dimension index) to keep track of the site associated with each row.
D=dataset(Y=dat['streamflow'],X=dat['nino'],iDim=dat['space_index'])
# Define formulas of the model - same as previously, but now some parameters will vary in space and hence be treated as processes
f=c('mu=m0+m1*nino','sigma=s0')
# Create a 'space' dimension, with lon-lat coordinates and the corresponding Haversine distance
space=dimension('space',coord=stations[c('lon','lat')],d=distance('Haversine'))
# m0 is a "free" spatial process - no hyperdistribution
m0=process('m0',dim=space,init=1) # no hyper-distribution by default
# m0 is a Gaussian iid spatial process
m1=process('m1',dim=space,init=0,dist='Gaussian_IID',
           par=list(parameter(name='m1_hypermean',init=0),
                    parameter(name='m1_hypersdev',init=1)))
# s0 is a parameter => it is assumed to be the same for all sites.
s0=parameter('s0',init=1,priorDist='FlatPrior+')
# Assemble the model - note the distinction between parameters and processes
mod=model(dataset=D,parentDist='LogNormal',par=list(s0),process=list(m0,m1),formula=f)
```

As previously, the function `STooDs` is called to perform MCMC sampling. 

```{r}
wk=file.path(getwd(),'man','readme','wk')
STooDs(model=mod,workspace=wk)
```

MCMC traces are shown below. Note that the processes appear in light green, and their hyper-parameters in dark green.

```{r fig.height=15, fig.width=9}
mcmc=readMCMC(file=file.path(wk,'MCMC.txt'))
plotMCMC.trace(mcmc,mod,panelPerCol=4)
```

The function `plotMCMC.par` now plots histograms for parameters and hyper-parameters.

```{r fig.height=3, fig.width=9}
plotMCMC.par(mcmc,mod)
```

The function `plotMCMC.process` can be used to see how the processes vary across the dimension (here, the sites). Values for the nino effect m1 are largely negative at all sites, confirming that El Nino has a consistent negative effect across this region.

```{r fig.height=6, fig.width=9}
plotMCMC.process(mcmc,mod)
```

# Going further

This README provides a very quick overview of the basic usage of the `RSTooDs` package. For more advanced usages, please consult the documentation of the functions and the vignettes of the package. In particular, the following capabilities of `RSTooDs` are potentially useful:

1. The use of several random variables.
2. The handling of censored data.
3. The use of non-iid spatial or temporal processes.
4. The use of nearest-neighbour processes for large datasets.
5. The use of identifiability constraints for 'hidden covariates' models (aka 'latent variables' or 'latent factors' models).
6. etc.

