---
title: "Intermittent spring river flows in Eastern Australia"
author: "Ben Renard"
date: "30/09/2019"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,results='hide')
```

# Preliminaries

Load RSTooDs:

```{r}
library(RSTooDs)
```

# Single-site analysis

## Selecting data
```{r}
oneSite <- data.frame(year=springFlow$year,flow=springFlow$station2)
plot(oneSite,pch=19,xlab='Year',ylab='Spring mean flow [m3/s]')
dat <- dataset(Y=oneSite$flow)
```

## Estimating a lognormal distribution 

Parameters:

```{r}
meanlog <- parameter(name='meanlog',init=0) 
sdlog <- parameter(name='sdlog',init=0.1) 
```

set Model and run!

```{r,results='markup'}
mod <- model(dataset=dat,parentDist='LogNormal',par=list(meanlog,sdlog),
             formula=c('meanlog = meanlog','sdlog = sdlog'))
STooDs(model=mod)
```

## Handling zeros, approach 1: modeling the occurence process

Create a dataset with two variables: a binary variable denoting the occurrence of a zero, and the flow for non-zero values.

```{r}
isZero=springFlow$station2==0 # zero-flow mask
v1 <- ifelse(isZero,1,0) # occurrence process
v2 <- springFlow$station2[!isZero] # non-zero flow values
dat <- dataset( # merge the two variables in a single dataset
  Y=c(v1,v2),
  var=c(rep('isZero',NROW(v1)),rep('nzFlow',NROW(v2)))
  )
```

### IID model

Parameters: need to add the probability of zero!
```{r}
p0 <- parameter(name='p0',init=0.2,priorDist='Uniform',priorPar=c(0,1)) 
meanlog <- parameter(name='meanlog',init=0) 
sdlog <- parameter(name='sdlog',init=0.1) 
```

set Model and run!

```{r}
mod <- model(dataset=dat,varName=c('isZero','nzFlow'),parentDist=c('Bernoulli','LogNormal'),
             par=list(p0,meanlog,sdlog),
             formula=c('prob = p0','meanlog = meanlog','sdlog = sdlog'))
STooDs(model=mod)
```


XXX
```{r}
mcmc <- read.table(file.path('STooDs_workspace','MCMC.txt'),header=T,sep=';')
npar=3;par(mfrow=c(npar,2))
for(i in 1:npar){
  plot(mcmc[,i],type='l',ylab=names(mcmc)[i])
  hist(mcmc[500:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE)
}
```

### ENSO effect

```{r}
dat <- dataset(
  Y=c(v1,v2),
  var=c(rep('isZero',NROW(v1)),rep('nzFlow',NROW(v2))),
  X=data.frame(nino=c(elNino$nino,elNino$nino[!isZero]))
  )
```

```{r}
p0 <- parameter(name='p0',init=-2) 
ENSO.p0 <- parameter(name='ENSO.p0',init=0) 
meanlog <- parameter(name='meanlog',init=0)
ENSO.meanlog <- parameter(name='ENSO.meanlog',init=0) 
sdlog <- parameter(name='sdlog',init=0.1) 
```

set Model and run!

```{r}
mod <- model(dataset=dat,
             varName=c('isZero','nzFlow'),parentDist=c('Bernoulli','LogNormal'),
             par=list(p0,ENSO.p0,meanlog,ENSO.meanlog,sdlog),
             formula=c(
               'prob = 1/(1+exp(-1*(p0+ENSO.p0*nino)))',
               'meanlog = meanlog+ENSO.meanlog*nino',
               'sdlog = sdlog'
               )
             )
STooDs(model=mod)
```

XXX
```{r,fig.height=10}
mcmc <- read.table(file.path('STooDs_workspace','MCMC.txt'),header=T,sep=';')
npar=5;par(mfrow=c(npar,2))
for(i in 1:npar){
  plot(mcmc[,i],type='l',ylab=names(mcmc)[i])
  hist(mcmc[500:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE)
}
```

## Handling zeros, approach 2: rectified distribution

### IID model 

Create a dataset with a single variable but treat zeros as censored values.

```{r}
isZero=springFlow$station2==0 # zero-flow mask
cType <- rep(0,NROW(springFlow)) # initialize censoring type at zero (no censoring)
cType[isZero] <- -1 # consider zeros as "less than" censored values 
# define STooDs dataset
dat <- dataset(Y=springFlow$station2,cType=cType)
```


Use a 3-parameter GPD distribution. Will be "chopped" below zero.

```{r}
theshold <- parameter(name='threshold',init=-0.1) 
scale <- parameter(name='scale',init=1) 
shape <- parameter(name='shape',init=0) 
```

set Model and run!

```{r}
mod <- model(dataset=dat,parentDist='GPD',
             par=list(theshold,scale,shape),
             formula=c('threshold = threshold','scale = scale','shape = shape'))
STooDs(model=mod)
```


XXX
```{r}
mcmc <- read.table(file.path('STooDs_workspace','MCMC.txt'),header=T,sep=';')
npar=3;par(mfrow=c(npar,2))
for(i in 1:npar){
  plot(mcmc[,i],type='l',ylab=names(mcmc)[i])
  hist(mcmc[1000:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE)
}
```

Translate this in terms of probability of zero. WARNING: STooDs convention for shape is opposite from R convention!

```{r}
require(evd)
p0=vector('numeric',NROW(mcmc))
for(i in 1:NROW(mcmc)){
  p0[i] <- pgpd(0,loc=mcmc$threshold[i],scale=mcmc$scale[i],shape=-1*mcmc$shape[i])
}
plot(p0,type='l',ylab='p0')
hist(p0[1000:NROW(p0)],main='',xlab='p0',freq=FALSE)
```

### ENSO effect
Just add ENSO to the previous dataset:

```{r}
dat <- dataset(Y=springFlow$station2,cType=cType,X=elNino['nino'])
```

Nino effect on the threshold parameter: will affect both the probability of zero and the distribution of non-zero flows.

```{r}
theshold <- parameter(name='threshold',init=-0.1)
ENSO.effect <- parameter(name='ENSO.effect',init=0) 
scale <- parameter(name='scale',init=1) 
shape <- parameter(name='shape',init=0) 
```

set Model and run!

```{r}
mod <- model(dataset=dat,parentDist='GPD',
             par=list(theshold,ENSO.effect,scale,shape),
             formula=c('threshold = threshold+ENSO.effect*nino',
                       'scale = scale','shape = shape'))
STooDs(model=mod)
```


XXX
```{r,fig.height=10}
mcmc <- read.table(file.path('STooDs_workspace','MCMC.txt'),header=T,sep=';')
npar=4;par(mfrow=c(npar,2))
for(i in 1:npar){
  plot(mcmc[,i],type='l',ylab=names(mcmc)[i])
  hist(mcmc[1000:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE)
}
```

## Comparison of the two approaches
2DO


<!-- # Multi-site analysis -->

<!-- We'll use the 'rectified distribution' approach. -->

<!-- ## Preparing data -->

<!-- ```{r} -->
<!-- nSite <- NCOL(springMean)-1 -->
<!-- # initialize empty data frames -->
<!-- Y=data.frame(value=numeric()) # flows -->
<!-- X=data.frame(nino_3.4=numeric()) # el nino -->
<!-- cType=data.frame(type=numeric()) # censoring type -->
<!-- iSpace=data.frame(space=integer()) # space index -->
<!-- for(site in 1:nSite){ -->
<!--   isZero=springMean[,site+1]==0 -->
<!--   Y=rbind(Y,data.frame(value=springMean[,site+1])) -->
<!--   X=rbind(X,data.frame(nino_3.4=elNino$nino_3.4)) -->
<!--   cType=rbind(cType,data.frame(type=ifelse(isZero,-1,0))) -->
<!--   iSpace=rbind(iSpace,data.frame(space=rep(site,NROW(springMean)))) -->
<!-- } -->

<!-- dat <- dataset(Y=Y,X=X,iDim=iSpace,cType=cType) -->

<!-- ``` -->

<!-- ## Model 1: everything is site-specific  -->
<!-- Space dimension and associated processes: -->

<!-- ```{r} -->
<!-- space <- dimension(name='space',coord=stations[c('lon','lat')],d=distance('Haversine'),z=stations['area']) -->
<!-- threshold <- process(name='threshold',dim=space,init=-2) -->
<!-- scale <- process(name='scale',dim=space,init=1) -->
<!-- shape <- process(name='shape',dim=space,init=0)  -->
<!-- ``` -->

<!-- set Model and run! -->

<!-- ```{r} -->
<!-- mod <- model(dataset=dat,parentDist='GPD', -->
<!--              process=list(threshold,scale,shape), -->
<!--              formula=c('threshold=scale*threshold','scale=scale','shape=shape')) -->
<!-- workspace <- file.path(wd,'STooDs_workspace') -->
<!-- STooDs(model=mod,workspace=workspace,dir.exe=dir.exe) -->
<!-- ``` -->

<!-- XXX -->
<!-- ```{r} -->
<!-- mcmc <- read.table(file.path(workspace,'MCMC.txt'),header=T,sep=';') -->
<!-- ``` -->


<!-- ```{r} -->
<!-- boxplot(mcmc[1000:NROW(mcmc),1:nSite],names=1:nSite, -->
<!--         xlab='Station',ylab='threshold') -->
<!-- ``` -->


<!-- ```{r} -->
<!-- boxplot(mcmc[1000:NROW(mcmc),nSite+(1:nSite)],names=1:nSite, -->
<!--         xlab='Station',ylab='scale') -->
<!-- `` -->
<!-- ```{r} -->
<!-- sc <- apply(mcmc[1000:NROW(mcmc),nSite+(1:nSite)],2,median) -->
<!-- plot(log(space$z$area),sc) -->
<!-- ```` -->


<!-- ```{r} -->
<!-- boxplot(mcmc[1000:NROW(mcmc),2*nSite+(1:nSite)],names=1:nSite, -->
<!--         xlab='Station',ylab='shape') -->
<!-- ``` -->


<!-- ## Model 1: Add some constraints  -->
<!-- Space dimension and associated processes: -->

<!-- ```{r} -->
<!-- # p0 IID Gaussian spatial process -->
<!-- p0.mean <- parameter(name='p0.mean',init=-2) -->
<!-- p0.sd <- parameter(name='p0.sd',init=0.1,priorDist='FlatPrior+') -->
<!-- p0 <- process(name='p0',dim=space,init=-2, -->
<!--               dist='Gaussian_IID', -->
<!--               par=list(p0.mean,p0.sd),iScalarPar=c(1,2)) -->
<!-- # ENSO effect on p0: constant -->
<!-- ENSO.p0 <- parameter(name='ENSO.p0',init=0) -->
<!-- # meanlog Gaussian spatial process -->
<!-- meanlog.mu0 <- parameter(name='meanlog.mu0',init=0) -->
<!-- meanlog.Aeffect <- parameter(name='meanlog.Aeffect',init=0) -->
<!-- meanlog.sill <- parameter(name='meanlog.sill',init=0.1,priorDist='FlatPrior+') -->
<!-- meanlog.range <- parameter(name='meanlog.range',init=1.,priorDist='Exponential',priorPar=c(0,5)) -->
<!-- meanlog <- process(name='meanlog',dim=space,init=0, -->
<!--                    dist='Gaussian', -->
<!--                    par=list(meanlog.mu0,meanlog.Aeffect,meanlog.sill,meanlog.range), -->
<!--                    fMean='meanlog.mu0+meanlog.Aeffect*area', -->
<!--                    fCovar='(meanlog.sill^2)*exp(-Dspace/meanlog.range)') -->
<!-- # ENSO effect on meanlog: constant -->
<!-- ENSO.meanlog <- parameter(name='ENSO.meanlog',init=0)  -->
<!-- # sdlog: no contraint -->
<!-- sdlog <- process(name='sdlog',dim=space,init=0.1)  -->
<!-- ``` -->

<!-- set Model and run! -->

<!-- ```{r} -->
<!-- mod <- model(dataset=dat,parentDist=c('Bernoulli','LogNormal'), -->
<!--              process=list(p0,meanlog,sdlog), -->
<!--              par=list(ENSO.p0,ENSO.meanlog), -->
<!--              formula=c( -->
<!--                'prob = 1/(1+exp(-1*(p0+ENSO.p0*nino_3.4)))', -->
<!--                'meanlog = meanlog+ENSO.meanlog*nino_3.4', -->
<!--                'sdlog = sdlog' -->
<!--                ) -->
<!--              ) -->
<!-- workspace <- file.path(wd,'STooDs_workspace') -->
<!-- STooDs(model=mod,workspace=workspace,dir.exe=dir.exe) -->
<!-- ``` -->

<!-- XXX -->
<!-- ```{r} -->
<!-- mcmc <- read.table(file.path(workspace,'MCMC.txt'),header=T,sep=';') -->
<!-- npar=2;par(mfrow=c(npar,2)) -->
<!-- for(i in 1:npar){ -->
<!--   plot(mcmc[,i],type='l',ylab=names(mcmc)[i]) -->
<!--   hist(mcmc[500:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE) -->
<!-- } -->
<!-- ``` -->

<!-- $p_0$ process: -->
<!-- ```{r} -->
<!-- boxplot(mcmc[500:NROW(mcmc),2+(1:nSite)],names=1:nSite, -->
<!--         xlab='Station',ylab='Probability of zero') -->
<!-- npar=2;par(mfrow=c(npar,2)) -->
<!-- for(i in 2+nSite+(1:npar)){ -->
<!--   plot(mcmc[,i],type='l',ylab=names(mcmc)[i]) -->
<!--   hist(mcmc[500:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE) -->
<!-- } -->
<!-- ``` -->

<!-- $meanlog$ process: -->
<!-- ```{r,fig.height=10} -->
<!-- boxplot(mcmc[500:NROW(mcmc),2+nSite+2+(1:nSite)],names=1:nSite, -->
<!--         xlab='Station',ylab='meanlog') -->
<!-- npar=4;par(mfrow=c(npar,2)) -->
<!-- for(i in 2+nSite+2+nSite+(1:npar)){ -->
<!--   plot(mcmc[,i],type='l',ylab=names(mcmc)[i]) -->
<!--   hist(mcmc[500:NROW(mcmc),i],main='',xlab=names(mcmc)[i],freq=FALSE) -->
<!-- } -->
<!-- ``` -->

<!-- $sdlog$ process: -->
<!-- ```{r} -->
<!-- boxplot(mcmc[500:NROW(mcmc),2+nSite+2+nSite+4+(1:nSite)],names=1:nSite, -->
<!--         xlab='Station',ylab='sdlog') -->
<!-- ``` -->
