---
title: "Examples of STooDs models"
subtitle: "Comparing models expressed as equations, STooDs code or graph"
author: "Ben Renard"
date: "30/03/2020"
output: 
  html_document: 
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Examples of STooDs models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction and aims
This vignette aims at providing a few examples of typical STooDs models. Each model is expressed in three ways: (1) mathematical equations; (2) translation of these equations as STooDs code; (3) representation of the resulting STooDs model as a graph. All models are based on the same dataset of East Australian floods

# Preliminaries

Start by loading RSTooDs:

```{r}
library(RSTooDs)
```

The dataset `springMax` used here described the spring maximum flow (in mm) at 42 sites in East Australia, over the period 1951-2014. As in previous case studies, the original "wide-format" dataset is transformed into a "long-format" vector format. We also add an extra column to keep track of the values taken by the ENSO index $x_t$, used as a covariate.

```{r}
nT <- NROW(springFlood); nS <- NCOL(springFlood)-1
longDataset <- data.frame(Qmax=numeric(nT*nS),nino=numeric(nT*nS),space=integer(nT*nS),time=integer(nT*nS))
for(s in 1:nS){
  indx=((s-1)*nT+1):(s*nT)
  longDataset[indx,]=cbind(Qmax=springMax[,s+1],nino=elNino$nino,space=rep(s,nT),time=1:nT)
}
```

Data from site 23 are plotted below as an illustration.

```{r}
plot(springMax$year,springMax$station23,pch=19,
     xlab='Year',ylab='Spring maximum flow [mm]')
```

# Single site: basic *iid* model

## In mathematical notation 

$$y_t \sim Gumbel(\varphi_{1,t},\varphi_{2,t})$$

$$
\begin{cases}
  \varphi_{1,t}=\mu_0 \\
  \varphi_{2,t}=\sigma_0 
\end{cases}
$$

## In STooDs 

The code below uses site 23 as an example.

```{r}
mask <- (longDataset$space==23)
dat <- dataset(Y=data.frame(Q=longDataset$Qmax[mask]))
mu0 <- parameter(name='mu0',init=20) 
sigma0 <- parameter(name='sigma0',init=5) 
mod <- model(dataset=dat,
             parentDist='Gumbel',
             par=list(mu0,sigma0),
             formula=c('location=mu0','scale=sigma0'))
```

## As a graph

```{r,results='asis',message=F, warning=F}
plot(mod)
```

# Single site: time-varying model

## In mathematical notation 

$$y_t \sim Gumbel(\varphi_{1,t},\varphi_{2,t})$$

$$
\begin{cases}
  \varphi_{1,t}=\mu_0(1+\mu_1x_t) \\
  \varphi_{2,t}=\sigma_0 
\end{cases}
$$

## In STooDs 

```{r}
mask <- (longDataset$space==23)
dat <- dataset(Y=data.frame(Q=longDataset$Qmax[mask]),
               X=data.frame(nino=longDataset$nino[mask]))
mu0 <- parameter(name='mu0',init=20) 
mu1 <- parameter(name='mu1',init=0)
sigma0 <- parameter(name='sigma0',init=5)
mod <- model(dataset=dat,
             parentDist='Gumbel',
             par=list(mu0,mu1,sigma0),
             formula=c('location=mu0*(1+mu1*nino)','scale=sigma0'))
```

## As a graph

```{r,results='asis',message=F,warning=F}
plot(mod)
```


# All sites: site-specific model

## In mathematical notation 

$$y_{t,s} \sim Gumbel(\varphi_{1,t,s},\varphi_{2,t,s})$$

$$
\begin{cases}
  \varphi_{1,t,s}=\mu_{s} \\
  \varphi_{2,t,s}=\sigma_{s} 
\end{cases}
$$

## In STooDs 

```{r}
dat <- dataset(Y=data.frame(Q=longDataset$Qmax),
               iDim=data.frame(space=longDataset$space))
space <- dimension(name='space',coord=OZstations[c('lon','lat')],d=distance('Haversine'))
mu <- process(name='mu',init=20,dim=space) 
sigma <- process(name='sigma',init=5,dim=space)
mod <- model(dataset=dat,
             parentDist='Gumbel',
             process=list(mu,sigma),
             formula=c('location=mu','scale=sigma'))
```

## As a graph

```{r,results='asis',message=F,warning=F,fig.height=8,fig.width=10}
plot(mod)
```

# All sites: constraining spatial variability (the Index Flood model)

## In mathematical notation 

$$y_{t,s} \sim Gumbel(\varphi_{1,t,s},\varphi_{2,t,s})$$

$$
\begin{cases}
  \varphi_{1,t,s}=\mu_{s} \\
  \varphi_{2,t,s}=\gamma \times \mu_{s} 
\end{cases}
$$

## In STooDs 

```{r}
dat <- dataset(Y=data.frame(Q=longDataset$Qmax),
               iDim=data.frame(space=longDataset$space))
space <- dimension(name='space',coord=OZstations[c('lon','lat')],d=distance('Haversine'))
mu <- process(name='mu',init=20,dim=space) 
gamma <- parameter(name='gamma',init=0.1,priorDist='LogNormal',priorPar=c(log(0.1),1))
mod <- model(dataset=dat,
             parentDist='Gumbel',
             par=gamma,process=mu,
             formula=c('location=mu','scale=gamma*mu'))
```

## As a graph

```{r,results='asis',message=F,warning=F,fig.height=8,fig.width=10}
plot(mod)
```


# All sites: a less constrained Index Flood model

## In mathematical notation 

$$y_{t,s} \sim Gumbel(\varphi_{1,t,s},\varphi_{2,t,s})$$

$$
\begin{cases}
  \varphi_{1,t,s}=\mu_{s} \\
  \varphi_{2,t,s}=\gamma_{s} \times \mu_{s} 
\end{cases}
$$

$$
\gamma_{s} \sim N(m_\gamma,s_\gamma)
$$

## In STooDs 

```{r}
dat <- dataset(Y=data.frame(Q=longDataset$Qmax),
               iDim=data.frame(space=longDataset$space))
space <- dimension(name='space',coord=OZstations[c('lon','lat')],d=distance('Haversine'))
mu <- process(name='mu',init=20,dim=space)
hypermean <- parameter(name='hypermean',init=0.1) 
hypersd <- parameter(name='hypersd',init=0.05) 
gamma <- process(name='gamma',init=0.1,dim=space,
                 dist='Gaussian_IID',par=list(hypermean,hypersd))
mod <- model(dataset=dat,
             parentDist='Gumbel',
             process=list(mu,gamma),
             formula=c('location=mu','scale=gamma*mu'))
```

## As a graph

```{r,results='asis',message=F,warning=F,fig.height=8,fig.width=10}
plot(mod)
```

